An Efficient Minibatch Acceptance Test for MCMC

### 2016-07-05 nips_2016.tex
fixed lower bound

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^N p(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^N$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $N$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the samples.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. This $P_a$ satisfies detailed balance, so if
one samples long enough, one will arrive at a stationary distribution matching the posterior, though
a burn-in period and/or only using samples at regular intervals is often done in practice.

Unfortunately, computing $f$ requires the use of all $N$ training data
points. Moreover, it is difficult to design tests using substantially
fewer than $N$ points that also satisfy detailed balance. To reconcile
these competing
objectives,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and tests the hypothesis that the sample $\theta'$ should be
accepted or rejected based on a probability ratio involving the
additional random variable $u\in [0,1]$. If the test cannot make a
decision with sufficient confidence, then the minibatch size is
increased and the test repeats. This process continues until a
decision. The bounds depend on either a central limit theorem
\cite{cutting_mh_2014} or a concentration bound
\cite{icml2014c1_bardenet14} requiring direct bounds on the log
likelihood ratio.  The problem with both methods is the expense of
resolving small differences.  In the worst case, all $N$ data points
may be needed, and as we show next this worst case can occur
about $\sqrt{N}$ times during the performance of $N$ tests.

We next prove that for some simple distributions, the lower
bound of average number of data instance consumed for one iteration of
~\cite{cutting_mh_2014,icml2014c1_bardenet14} is
$\Omega(\sqrt{N})$, where $N$ is the number of data points, see
Lemma~\ref{lem:efficienty_cutting}.

\begin{lemma}\label{lem:efficienty_cutting}
If the sample mean $\bar{l}$ has uniform distribution on $[-L,L]$,
the lower bound of expected data consumed of adaptive sampling method
for one iteration is $\frac{\eta s_{l}Q}{2L}\sqrt{N}$, where $\bar{l}
= \frac{1}{n}\sum_{i=1}^{n} \left (\log p(x_i; \theta') - \log p(x_i|
\theta_t)\right )$, $\eta = \phi^{-1}_{n-1}(1-\varepsilon)$,
$Q=\frac{p(\theta_t)q(\theta'|\theta_t)}{p(\theta')q(\theta_t|\theta')}$,
$s_l$ is the population standard deviation of $l_i$, $\phi_{n-1}()$ is
the CDF of t distribution with degree of freedom $n-1$, and $n$ is the
minibatch size.
\end{lemma}

\begin{proof} see appendix
  \end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

\section{Theoretical Results}\label{sec:theory}

In this section, we firstly explore the convergence properties of our MH test. Secondly, we compare the efficiency between our MH test and adaptive MH test. Due to space constraints, all
our proofs are in Appendix~\ref{app:proofs}.

We first introduce some notation. Let our true and approximated acceptance probabilities be $P_a =
g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} > 0)$.  Define $X_{\xi} =
\varepsilon - X_{\rm norm}$ which means, invoking Equations~\ref{eq:relationship}
and~\ref{eq:deconvolution}, that $X_\xi$ is the random variable characterizing the accuracy of our
$X_{\rm norm}$.  Thus, $\Delta' = \Delta + X_{\rm norm} + X_\xi$. We assume the standard deviation
of $X_{\rm norm}$ is less than $\pi/\sqrt{3}$, so that we can use Equation~\ref{eq:deconvolution}.
We denote $F_X(x)$ as the CDF of $X$.

>>>>

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^N p(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^N$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $N$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta',\theta_t) = \min\left\{ 1, \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)}
\right\}
\end{equation}
and with probability $\alpha(\theta',\theta_t)$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta',\theta_t)$

The acceptance test guarantees detailed balance, so that
\begin{equation}
  p(\theta)q(\theta',\theta)\alpha(\theta',\theta) = p(\theta')q(\theta,\theta')\alpha(\theta,\theta')
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta',\theta_t) = 
\min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid\theta_t)} \right\},
\end{equation}
where $p(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points. Moreover, it is difficult to
design tests using substantially fewer than $N$ points that also
satisfy detailed balance. To reconcile these competing
objectives,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta',\theta_t)$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\sqrt{N}$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta',\theta_t)$ in
the equivalent form $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ where
\begin{equation}
\Lambda(\theta,\theta') = \frac{1}{N}\sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \frac{1}{N}\log\left(u\frac{q(\theta'|\theta)p(\theta)}{q(\theta|\theta')p(\theta')}\right)
\end{equation}

By the central limit theorem, we expect $\Lambda(\theta_t,\theta')$ to be approximately
Normal. To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:

\begin{equation}
\Lambda^*(\theta,\theta') = \frac{1}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

The acceptance test then becomes a statistical test of the hypothesis
that $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ by
establishing that $\Lambda^*(\theta,\theta')$ is substantially larger
than $\psi(u,\theta,\theta')$.  In ~\cite{cutting_mh_2014} an
asymptotic central limit argument was used to derive this gap, while
in \cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

We next show that for some simple distributions, the lower
bound of average number of data instance consumed for one iteration of
~\cite{cutting_mh_2014} and ~\cite{icml2014c1_bardenet14} is
$\Omega(\sqrt{N})$, where $N$ is the number of data points.

Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta_t)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta_t)^2\right) =
  (\theta'-\theta_t)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta_t-\frac{\theta'-\theta_t}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta_t)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta_t)$
and $\psi(u,\theta',\theta_t)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\mu,1/N)$, so we take $q(\theta'\mid\theta_t) =
\phi((\theta'-\theta_t)N)$, where $\phi(x)$ is a Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta',\theta_t)=q(\theta_t,\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta_t)$ reduces to $1/N\log
u$.

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta_t, \theta', u)$, the
  test from \cite{icml2014c1_bardenet14} consumes all $N$ samples. 
\end{lemma}

\begin{proof} Assume $(\theta' - \theta_t) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and
  $\theta_t -0.5 \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events
  occur with probability $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and $p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively. 
  Choose $u_0$
  to satisfy
  \begin{equation}
    \log u_0 = -N(\theta'-\theta_t)\left(\theta_t-0.5+\frac{\theta'-\theta_t}{2}\right) 
  \end{equation}
  and rewrite the test now as
  \begin{equation}
\frac{1}{b}\sum_{i=1}^b x_i-\left(\theta_t+\frac{\theta'-\theta_t}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) > 0
  \end{equation}
  The above choice of $u_0$ ensures that the terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the
  test will never correctly succeed.

  Furthermore, if we sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be sufficiently different
  from 0.5 to allow the test to succeed.

 The choices above for $\theta_t$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-1,0.375]
  \end{equation}
  the size of this range is at least $\exp([-2.5,-1.125]) = [0082,0.286]$ and occurs
  with probability at least $p_2=0.2044$. The samples $\theta_t$, $(\theta'-\theta_t)$
  and $u$ are drawn independently and so the total probability of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented

proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

\section{Analysis}\label{sec:theory}

In this section, we first perform an error analysis for our new
acceptance test. Second, we follow the implications of this analysis for accuracy of
the estimated target distribution. Third, we compare the speed and accuracy of
our test with prior work. Due to space constraints, all our proofs
are in Appendix~\ref{app:proofs}.

We first introduce some notation. Let our true (full dataset) and
approximate (minibatch) acceptance probabilities be $P_a = g(\Delta) =
\Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} > 0)$.
Define $X_{\xi} = \varepsilon - X_{\rm norm}$ which means, invoking
Equations~\ref{eq:relationship} and~\ref{eq:deconvolution}, that
$X_\xi$ is the random variable characterizing the accuracy of our
$X_{\rm norm}$.  Thus, $\Delta' = \Delta + X_{\rm norm} + X_\xi$. We
assume the standard deviation of $X_{\rm norm}$ is less than
$\pi/\sqrt{3}$, so that we can use Equation~\ref{eq:deconvolution}.
We denote $F_X(x)$ as the CDF of $X$.
